Scoring Patients

### 2017-04-04 Patient Report.Rmd
Adding new analyses

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient-OLD', sep='/')

@@

patient.info = read.patient.info(patient.file)

patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = subset(sum.patient.data, Initial.Analysis)

patient.info = subset(patient.info, Patient %in% sum.patient.data$Patient & Batch.Name %in% levels(patient.info$Batch.Name)[1:5])

@@

@@

      if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology)
        }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

@@

The result looks like is two data matrices:

@@

```{r general, warning=F, echo=F, message=F}
gen = do.call(rbind, lapply(patient.data, function(x) nrow(x[['seg.vals']])))
colnames(gen) = c('Number.Regions')

@@

pander(sum.patient.data[,c('Patient','Status','years','total.samples','highest.path','Number.Regions')], justify='left')
```

# Segmented Value Distributions
```{r echo=F, message=F, warning=F, fig.height=15, fig.width=10}

@@

```{r var_reg, echo=F, message=F, warning=F}
rm(prog.regions, np.regions)
for (patient.name in sum.patient.data$Patient) {
  patient = patient.data[[patient.name]]

  sum.patient.data[which(sum.patient.data$Patient == patient.name), 'var.regions'] = nrow(patient$variable.regions)

  if (subset(sum.patient.data, Patient == patient.name)$Status == 'P') {
    if (!exists('prog.regions')) {
      prog.regions = patient$variable.regions[,c('chrom','start.pos','end.pos')]
    } else {
      prog.regions = rbind(prog.regions, patient$variable.regions[,c('chrom','start.pos','end.pos')])
    }
  } else {
    if (!exists('np.regions')) {
      np.regions = patient$variable.regions[,c('chrom','start.pos','end.pos')]
    } else {
      np.regions = rbind(np.regions, patient$variable.regions[,c('chrom','start.pos','end.pos')])
    }
  }
}

pander(sum.patient.data %>% group_by(Status) %>% summarise(mean.num.regions = mean(var.regions), sd.num.regions = sd(var.regions)), justify='left')

prog.regions$length = prog.regions$end.pos - prog.regions$start.pos
np.regions$length = np.regions$end.pos - np.regions$start.pos

chrs = read.table('hg19_chrom.txt',sep='\t', header=T)
prog.regions = merge(prog.regions, chrs, by.x='chrom', by.y='chr') 
np.regions = merge(np.regions, chrs, by.x='chrom', by.y='chr')

prog.regions$chr.ratio = prog.regions$length.x/prog.regions$length.y
np.regions$chr.ratio = np.regions$length.x/np.regions$length.y

ttR = t.test(prog.regions$chr.ratio, np.regions$chr.ratio)

np = np.regions %>%
  group_by(chrom) %>%
  summarise(lengths = sum(length.x))

prog = prog.regions %>%
  group_by(chrom) %>%
  summarise(lengths = sum(length.x))

m = merge(prog, np, by='chrom', all.x=T)
tt = t.test(m$lengths.x, m$lengths.y)

plot(hclust( dist(chrs$length)), main='Chromosome cluster by length')

## Grouping the chromosomes by length 1-12, 13-22

#t.test(subset(np.regions, chrom %in% c(1:12))$chr.ratio, subset(np.regions, chrom %in% c(13:22))$chr.ratio)

@@

df = data.frame( 'short' = rbind( 'np' = sum(table(np.regions$chrom)[1:12]),
       'p' = sum(table(prog.regions$chrom)[1:12])),
                 'long' = rbind( 'np' = sum(table(np.regions$chrom)[13:22], na.rm=T),
       'p' = sum(table(prog.regions$chrom)[13:22], na.rm=T)))

ftL = fisher.test(df)
```

So it appears that adjusting for chromosome length (looking at the ratio of the chromosome rather than the raw bp) there is still a significant difference between progressers and non: t-test p-value `r ttR$p.value`

There appears to be no enrichment based on the length of chromosomes either (fisher `r ftL$p.value`)

```{r cor, echo=F, message=F, warning=F, fig.height=10,fig.width=10, fig.align='left'}

@@

cors = list()
for (name in names(patient.data)) {

  sample.normalised.segvals = patient.data[[name]]$norm.seg.vals # prog
  sample.normalised.segvals =  sample.normalised.segvals[, intersect(patient.data[[name]]$info$Samplename, colnames(sample.normalised.segvals)) ]

@@

diffPNP = wilcox.test(unlist(lapply(cors[subset(sum.patient.data, Status == 'P')$Patient], function(x) sum(x$diff))),
            unlist(lapply(cors[subset(sum.patient.data, Status == 'NP')$Patient], function(x) sum(x$diff))))
```

There's no clear difference (currently) between P and NP for the similar correlation (`r simPNP$p.value`) or the inverse (`r diffPNP$p.value`) correlations.  There are other ways we may be able to look at this still.

@@

rm(var.regions)
for (patient.name in names(patient.data)) {
  print(patient.name)
  regions = patient.data[[patient.name]]$variable.regions[,c('chrom','arm','start.pos','end.pos')]
  if (nrow(regions) <= 0) next

  reg = cancer.consensus.genes(regions, ccgenes)
  x = patient.data[[patient.name]]$variable.regions[which(!is.na(reg$Gene.Symbols)), patient.data[[patient.name]]$info$Samplename]

@@

gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
m = melt(gc, id.vars=c('Gene.Symbols'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')

@@

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  print(name)
  sn = grep('HGD|IMC', patient.data[[name]]$info$Pathology)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next
  samplenames = patient.data[[name]]$info$Samplename

  ## HGD
  pvnp[name, 'hgd'] = length(which(patient.data[[name]]$variable.regions[samplenames[sn]] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[samplenames[sn]] <= loss.threshold))

  # BE-LGD
  pvnp[name, 'be'] = length(which(patient.data[[name]]$variable.regions[samplenames[-sn]] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[samplenames[-sn]] <= loss.threshold))
}

# No difference

@@

@@

pander(pvnp, justify='left', 
       caption=paste("Variable regions found in HGD vs non-HGD samples of progessor patients.  There's no significant difference between them (t-test p=",signif(tt$p.value,2),').  Not all patients have HGD samples (NAs). There is also no significant correlation between the number/ratio of variable regions found in HGD samples vs non (Pearsons p=',signif(ct$p.value, 2),')',  sep=''))

@@

>>>>

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient', sep='/')

@@

patient.info = read.patient.info(patient.file)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

@@

sum.patient.data = summarise.patient.info(patient.info)

#patient.info = subset(patient.info, Patient %in% sum.patient.data$Patient & Batch.Name %in% levels(patient.info$Batch.Name)[1:5])

@@

  missing = c(setdiff(colnames(segvals)[-(1:5)], subset(patient.info, Patient == patient.name)$Samplename),
                    setdiff(subset(patient.info, Patient == patient.name)$Samplename, colnames(segvals)[-(1:5)]))

  if (length(missing) > 0) 
    warning(paste("Missing", length(missing), "sample(s) in patient", patient.name))

@@

      if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology, Samplename)
        }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

## Check that all of the samples are there
patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year,Samplename)
pt = 'AH0254'
info = subset(patient.info, Patient == 'AH0254')
lapply(names(patient.data), function(pt) {
  missing = which(!patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals))
  if (length(missing) > 0) {
    msg = paste(pt,": Missing samples - ", paste(patient.data[[pt]]$info$Samplename[missing], collapse=', ') , sep='')
    msg = paste(msg, "\nFrom batch: ", paste(unique(patient.data[[pt]]$info$Batch.Name[missing]), collapse=', '))
    warning(msg)
  }
})

save(patient.data, file='patients.Rdata')

@@

The result looks like two data matrices:

@@

```{r general, warning=F, echo=F, message=F}
gen = do.call(rbind, lapply(patient.data, function(df) {
  cbind('n.segments'=nrow(df[['seg.vals']]), 
        'median.seg.len'=median(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]),
        'mean.seg.len'=mean(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]), 
        'sd.seg.len'=sd(df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')]))
}))
rownames(gen)  = names(patient.data)

@@

pander(sum.patient.data[,c('Patient','Status','years','total.samples','highest.path','n.segments', 'median.seg.len')], justify='left')

#seg.lens = lapply(patient.data, function(df) df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')] )

#wilcox.test(log(unlist(seg.lens[subset(sum.patient.data, Status == 'P')$Patient])),
#log(unlist(seg.lens[subset(sum.patient.data, Status == 'NP')$Patient])))

#hist(log(unlist(seg.lens[subset(sum.patient.data, Status == 'P')$Patient])))
#hist(log(unlist(seg.lens[subset(sum.patient.data, Status == 'NP')$Patient])))

ggplot(sum.patient.data) + geom_boxplot( aes(y=median.seg.len, x=Status, group=Status, fill=Status) ) +
  labs(title="Median segment length by patient status")

There's no significant difference between the length of segments in progressors vs non.  There's a larger spread of lengths in progressors, but the means are nearly identical.

# Segmented Value Distributions
```{r echo=F, message=F, warning=F, fig.height=15, fig.width=10, eval=F}

@@

```{r var_reg, echo=F, message=F, warning=F}
prog.pt = subset(sum.patient.data, Status == 'P')$Patient
np.pt = subset(sum.patient.data, Status == 'NP')$Patient

var.seg.lengths = lapply(patient.data, function(df) {
  cbind('chr'=df$variable.regions$chrom, 'seg.len'=df$variable.regions$end.pos - df$variable.regions$start.pos)
})

segment.lengths = lapply(patient.data, function(df) {
  cbind('chr'=df$seg.vals$chrom, 'seg.len'=df$seg.vals$end.pos - df$seg.vals$start.pos)
})

prog.regions = sapply(var.seg.lengths[sum.patient.data$Patient], nrow)

sum.patient.data$Num.Var.Regions = as.vector(sapply(var.seg.lengths[sum.patient.data$Patient], nrow))

chrs = read.table('hg19_chrom.txt',sep='\t', header=T)
var.seg.lengths = lapply(var.seg.lengths, function(x) merge(x, chrs, by.x='chr', by.y='chr') )

segment.lengths= lapply(segment.lengths, function(x) merge(x, chrs, by.x='chr', by.y='chr') )

opcov = sapply(segment.lengths[prog.pt], function(x) {
  sum(as.numeric(x[['seg.len']]))/sum(as.numeric(chrs$length))
})

onpcov = sapply(segment.lengths[np.pt], function(x) {
  sum(as.numeric(x[['seg.len']]))/sum(as.numeric(chrs$length))
})

chr.ratios = lapply(var.seg.lengths, function(x) x[['seg.len']]/x[['length']])
# Segment coverage, variable regions
ttR = t.test(unlist(chr.ratios[subset(sum.patient.data, Status == 'P')$Patient]), 
             unlist(chr.ratios[subset(sum.patient.data, Status == 'NP')$Patient]))

var.seg.lengths = lapply(var.seg.lengths, function(df) {
  as.data.frame(df %>%
  group_by(chr, length) %>%
  summarise(seg.lengths = sum(seg.len), 
            chr.ratio = sum(seg.len)/length[1]))
})

pcov = median(sapply(var.seg.lengths[prog.pt], function(x) 
  sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)

npcov = median(sapply(var.seg.lengths[np.pt], function(x) 
  sum(as.numeric(x[['seg.lengths']]))/sum(as.numeric(x[['length']])) ), na.rm=T)

## Grouping the chromosomes by length 1-12, 13-22
#t.test(subset(np.regions, chrom %in% c(1:12))$chr.ratio, subset(np.regions, chrom %in% c(13:22))$chr.ratio)

@@

#There may be an enrichment for segments in the shorter chromosomes amongst progressors but it's weak (fisher `r ftL$p.value`)
#df = data.frame( 'short' = rbind( 'np' = sum(table(np.regions$chrom)[1:12]),
#       'p' = sum(table(prog.regions$chrom)[1:12])),
#                 'long' = rbind( 'np' = sum(table(np.regions$chrom)[13:22], na.rm=T),
#       'p' = sum(table(prog.regions$chrom)[13:22], na.rm=T)))

#ftL = fisher.test(df)
```

So it appears that adjusting for chromosome length (looking at the ratio of the chromosome covered by segments rather than the raw bp) a significant difference in the coverage of the genome by CN segments that are variable between progressors and non: t-test p-value `r signif(ttR$p.value, 2)`.  Overall bp coverage of the sWGS is `r round(mean(c(opcov, onpcov)), 2)*100`%  (P `r round(mean(opcov),2)`, NP `r round(mean(onpcov), 2)`).

Variable regions cover `r round(pcov, 2)*100`% of the the progressors genome, and `r round(npcov, 2)*100`% of non-progressor genomes.

```{r cor, echo=F, message=F, warning=F, fig.height=10,fig.width=10, fig.align='left', eval=F}

@@

cors = list()
for (name in names(patient.data)) {
  sample.normalised.segvals = patient.data[[name]]$norm.seg.vals # prog
  sample.normalised.segvals =  sample.normalised.segvals[, intersect(patient.data[[name]]$info$Samplename, colnames(sample.normalised.segvals)) ]

@@

diffPNP = wilcox.test(unlist(lapply(cors[subset(sum.patient.data, Status == 'P')$Patient], function(x) sum(x$diff))),
            unlist(lapply(cors[subset(sum.patient.data, Status == 'NP')$Patient], function(x) sum(x$diff))))

#There's no clear difference (currently) between P and NP for the similar correlation (`r simPNP$p.value`) or the inverse (`r diffPNP$p.value`) correlations.  There are other ways we may be able to look at this still.

@@

rm(var.regions)
for (patient.name in names(patient.data)) {
  regions = patient.data[[patient.name]]$variable.regions[,c('chrom','arm','start.pos','end.pos')]
  if (nrow(regions) <= 0) next

  reg = cancer.consensus.genes(regions, ccgenes)
  x = patient.data[[patient.name]]$variable.regions[which(!is.na(reg$Gene.Symbols)), intersect(patient.data[[patient.name]]$info$Samplename, colnames(patient.data[[patient.name]]$variable.regions))  ]

@@

#gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
#gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
#m = melt(gc, id.vars=c('Gene.Symbols'))
#m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
#ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
#  geom_tile(aes(fill=value), color='white') + 
#  scale_fill_gradient('n patients',low='white',high='royalblue4') + labs(x="",y='CGC Genes')
```

@@

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  sn = which(colnames(patient.data[[name]]$variable.regions)[-(1:5)] %in% subset(patient.data[[name]]$info, grepl('HGD|IMC', Pathology))$Samplename)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next

  ## HGD
  pvnp[name, 'hgd'] = length(which(patient.data[[name]]$variable.regions[,sn] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[,sn] <= loss.threshold))

  # BE-LGD
  pvnp[name, 'be'] = length(which(patient.data[[name]]$variable.regions[-sn] >= gain.threshold | 
                                patient.data[[name]]$variable.regions[-sn] <= loss.threshold))
}

## TODO Check enrichment of variable regions found in BE vs HGD

@@

#fisher.test(p, simulate.p.value = T)

@@

pander(pvnp, justify='left', 
       caption=paste("Number of variable regions found in HGD vs non-HGD samples of progessor patients.  There's no significant difference between them (t-test p=",signif(tt$p.value,2),').  Not all patients have HGD samples (NAs). There is also no significant inverse correlation (r= ',signif(ct$estimate, 2),') between the number/ratio of variable regions found in HGD samples vs non (Pearsons p=',signif(ct$p.value, 2),')',  sep=''))

@@

This is just looking at the distribution of values per sample. The average ploidy per sample is around 1.  Currently we're calling variable regions based on this.
```{r, warning=F}
ks = list()
for (pt in names(patient.data)) {
  cols = intersect(arrange(patient.data[[pt]]$info, Pathology, Endoscopy.Year)$Samplename, colnames(patient.data[[pt]]$seg.vals[,-(1:5)]))
  segvals = patient.data[[pt]]$seg.vals[,cols]

  sdx = round(apply(segvals, 2, sd), 3)
  se = round(apply(segvals, 2, function(x) sd(x)/sqrt(length(x))), 3)

 gg = ggplot(melt(segvals)) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram(bins=10, fill='blue', color='grey',aes(value, ..density..)) +
        geom_density(aes(value+sdx/2), color='red', size=1) +
        geom_density(aes(value-sdx/2), color='green', size=1) +
        labs(title=paste(pt, '  (',patient.data[[pt]]$info$Status[1],')', sep=''))
  print(gg)   
  #plots[[pt]] = gg

  ks[[pt]] = sapply(segvals, function(x) ks.test(x+sdx, x-sdx)$statistic  )
  print(pt)
  print(colSums(abs(round(patient.data[[pt]]$seg.vals[,-(1:5)], 1)-1)))
}
